Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-01 Documents/Paper/ms.tex
affil
objects --> components

\title{Inference for Trans-dimensional Bayesian Models with Diffusive Nested
Sampling}
\author{Brendon J. Brewer}

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
objects in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. A common statistical example is finite mixture
modelling. In the Bayesian framework, these problems are typically solved using
one of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.
Both of the examples demonstrate phase transitions in the relationship between
the likelihood and the cumulative prior mass.}

\section{Introduction}\label{sec:introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Consider the following class of inference problems. There is some unknown
number, $N$, of objects in a (physical or metaphorical) region. Each of the $N$ objects has a property
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a two-dimensional position and a mass). These problems are often
challenging to solve due to the unknown (and potentially large) dimensionality.
In addition, they also have the so-called ``label-switching degeneracy'' issue
\citep{label_switching}, where the meaning of the model is invariant to
switching the identities of the objects. Finite mixture modelling is a common
statistical example.

To carry out the inference, we must assign a prior to the objects' properties
$\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.
The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

@@

The computational method used to calculate properties of the posterior distribution
is to generate samples from it using Markov Chain Monte Carlo (MCMC). Since
$N$ is unknown, an MCMC sampler needs to be able to jump between candidate
solutions with different numbers of objects, using either birth-death MCMC
\citep{birthdeath} or the reversible jump framework of \citet{rjmcmc}.

@@

\begin{figure}
\begin{center}
\includegraphics{pgm.pdf}
\caption{\it A probabilistic graphical model (PGM) depicting the kind
of model discussed in this paper. Produced using daft ({\tt daft-pgm.org}).
The number, $N$, of objects in the model is an unknown parameter, as are the
properties of the objects. The data depends on the properties of the objects.
Note that it is possible (and common) to have other parameters in the model
that point directly to the data. However, we have omitted such parameters
for simplicity.
\label{fig:pgm}}
\end{center}
\end{figure}

@@

% Consider terminology: "objects" vs. "components"

@@

If $N' > N$, i.e. the proposal is to add objects to the model,
the extra parameters needed, $\{\xx_i\}_{i=N+1}^{N'}$,
are drawn from their prior conditional on the current value of the
hyperparameters, i.e. the interim prior $p(\xx | \hyperparams)$.
If $N' < N$, i.e. the proposal is to remove objects from the model,
then $N - N'$ objects must be selected for removal. All
of the objects have the same probability of being selected for removal.

\subsection{Proposals that modify objects}\label{sec:proposal2}
We now consider a proposal distribution that modifies one or more of the
objects $\{\xx_i\}$, while keeping the number of objects $N$, as well as the
hyperparameters $\hyperparams$, fixed.

@@

Choosing just one object to change is most appropriate when the DNS
distribution is very constrained. When it is close to the prior, bolder
proposals that change more than one object at a time are possible. Hence,
we choose the number of objects to change from a heavy tailed distribution
wherer the most probable value is 1 but there is also a nontrivial probability
of proposing to change $\sim N$ objects at once.

\subsection{Proposals that change the hyperparameters,
keeping the objects fixed}\label{sec:proposal3}
Another kind of proposal that we will consider is a proposal that keeps all of
the objects fixed in place (i.e. leaves $\{\xx_i\}$) unchanged, but changes
the hyperparameter(s) from their current value $\hyperparams$
to a new value $\hyperparams'$. The proposal for the hyperparameter(s) is chosen
to satisfy detailed balance with respect to $p(\hyperparams)$ and should be
heavy-tailed.

The overall Metropolis acceptance probability
for this kind of move, if sampling the prior (or the constrained prior of DNS)
must include the following factor:
\begin{eqnarray}
\frac{\prod_{i=1}^N p(\xx_i | \hyperparams')}{\prod_{i=1}^N p(\xx_i | \hyperparams)}
\label{eqn:acceptance_prob}
\end{eqnarray}
Since this proposal leaves the objects $\{\xx_i\}$ fixed, it will usually not
affect the value of the likelihood, and therefore the likelihood will not need
to be recomputed. Therefore this proposal is most useful for mixing the values
of the hyperparameters $\hyperparams$ when the target distribution is highly
constrained.

\subsection{Proposals that change the hyperparameters
and all of the objects}\label{sec:proposal4}
The above proposals allow for changes to $N$, $\hyperparams$, and $\{\xx_i\}$,
and are therefore sufficient to allow for ``correct'' exploration of the
prior distribution, and the constrained prior distributions used in DNS.
However, they do not necessarily allow for {\it efficient} exploration, even
of the prior itself. The main reason for this is the inability for large
changes to be made to the hyperparameters $\hyperparams$. If the proposed change
from $\hyperparams$ to $\hyperparams'$ is large, the ratio in
Equation~\ref{eqn:acceptance_prob} is likely to be very small, and the move
will probably be rejected.

Therefore, we allow an additional move that changes $\hyperparams$ to a new
value $\hyperparams'$, but rather than leaving the objects $\{\xx_i\}$ fixed,
they are ``dragged''
so they represent the distribution $p(\xx | \hyperparams')$ rather than
$p(\xx|\hyperparams)$. We do this by making use of the transformation functions
$F(x; \hyperparams)$ and $G(x; \hyperparams)$ defined in
Section~\ref{sec:proposal2}.

The ``dragging'' process works as follows, and must be carried out on
each object:
\begin{eqnarray}
u_i &:=& F(\xx_i; \hyperparams)\\
\xx_i' &:=& G(\xx_i; \hyperparams')
\end{eqnarray}
In other words, the objects' parameters are transformed to $[0,1]$ using the
current value of the hyperparameters, and then transformed back using the
proposed value of the hyperparameters so they represent the new interim prior
rather than the old one.

@@

%\section{``Transit'' Example}
%\section{``Asteroseismology'' Example}
\section{``Galaxy Field'' Example}
Source detection is an important problem in astrophysics. Given some data,
usually one or more noisy, blurred, images of the sky, we would like to know
how many objects $N$ (such as stars or galaxies) are in the image, and their
properties (such as flux, size, orientation). Various approaches exist,
covering a wide spectrum between ad-hoc approaches and principled inference
approaches
\citep[e.g.][]{irwin, sextractor, dolphot, 2003MNRAS.338..765H, starfield}.
The more principled techniques tend to be much more computationally
intensive, so should only be applied to interesting or
especially challenging subsets of astronomical imaging. In fact, the approach
used by \citet{starfield} was essentially an early version of the methodology
presented in this paper.

In this section we apply
the DNS approach to a toy version of the problem of detecting and quantifying
extended objects such as galaxies in a noisy image.
We model each galaxy as a mixture of two concentric elliptical gaussian
components, with
the following eight parameters. Firstly, two
parameters $x_c$ and $y_c$ describe the central position of the galaxy within
the image. The flux $f$ describes the total integral of the galaxy's
intensity profile. The axis ratio $q$ describes the ellipticity of the galaxy
and $\theta$ its orientation angle with respect to horizontal. The radius of the
bigger gaussian is a parameter $w$.
Finally
we include a ``radius ratio'' $u$ describing the ratio of the radius of the
smaller gaussian with respect to that of the bigger gaussian, and a parameter
$v$ describing the fraction of the total flux in the smaller gaussian
(therefore the fraction of the light in the bigger gaussian is $1-v$).
In a coordinate system aligned with the major axis of the ellipse, the
surface brightness profile of a ``galaxy'' is given by
\begin{eqnarray}
\rho(x, y) &=&
\frac{f(1-v)}{2\pi w^2}\exp\left[-\frac{1}{2w^2}\left(qx^2 + y^2/q\right)\right]\\
&&+\frac{fv}{2\pi (uw)^2}\exp\left[-\frac{1}{2(uw)^2}\left(qx^2 + y^2/q\right)\right].
\end{eqnarray}
We generated a simulated $200 \times 200$ pixel noisy image
(Figure~\ref{fig:galaxyfield_data}) containing 47 ``galaxies'',
in order to test the algorithm.

@@

However, many of the proposal distributions used involve changing a subset
of the parameters, while keeping others fixed. Considerable speedups can be
achieved by, when appropriate, updating the mock signal to reflect the proposed
change to the model, rather than computing the entire mock signal from scratch.
For example, if the proposal is to add two new objects to the model, the
mock signal can simply be updated by adding the effect of the two new objects
to the model.

>>>>

\title{Inference for Trans-dimensional Bayesian Models with Diffusive Nested
Sampling}
\author{Brendon J. Brewer}
\date{Department of Statistics, The University of Auckland\\
Private Bag 92019, Auckland 1142, New Zealand\\
\vspace{0.3cm}
{\tt bj.brewer@auckland.ac.nz}}

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
components in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. A common statistical example is finite mixture
modelling. In the Bayesian framework, these problems are typically solved using
one of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.
Both of the examples demonstrate phase transitions in the relationship between
the likelihood and the cumulative prior mass.}

\section{Introduction}\label{sec:introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Consider the following class of inference problems. There is some unknown
number, $N$, of components in a (physical or metaphorical) region. Each of the $N$ components has a property
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a two-dimensional position and a mass). These problems are often
challenging to solve due to the unknown (and potentially large) dimensionality.
In addition, they also have the so-called ``label-switching degeneracy'' issue
\citep{label_switching}, where the meaning of the model is invariant to
switching the identities of the components. Finite mixture modelling is a common
statistical example.

To carry out the inference, we must assign a prior to the components' properties
$\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.
The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

@@

The computational method used to calculate properties of the posterior distribution
is to generate samples from it using Markov Chain Monte Carlo (MCMC). Since
$N$ is unknown, an MCMC sampler needs to be able to jump between candidate
solutions with different numbers of components, using either birth-death MCMC
\citep{birthdeath} or the reversible jump framework of \citet{rjmcmc}.

@@

\begin{figure}
\begin{center}
\includegraphics{pgm.pdf}
\caption{\it A probabilistic graphical model (PGM) depicting the kind
of model discussed in this paper. Produced using daft ({\tt daft-pgm.org}).
The number, $N$, of components in the model is an unknown parameter, as are the
properties of the components. The data depends on the properties of the components.
Note that it is possible (and common) to have other parameters in the model
that point directly to the data. However, we have omitted such parameters
for simplicity.
\label{fig:pgm}}
\end{center}
\end{figure}

@@

% Consider terminology: "components" vs. "components"

@@

If $N' > N$, i.e. the proposal is to add components to the model,
the extra parameters needed, $\{\xx_i\}_{i=N+1}^{N'}$,
are drawn from their prior conditional on the current value of the
hyperparameters, i.e. the interim prior $p(\xx | \hyperparams)$.
If $N' < N$, i.e. the proposal is to remove components from the model,
then $N - N'$ components must be selected for removal. All
of the components have the same probability of being selected for removal.

\subsection{Proposals that modify components}\label{sec:proposal2}
We now consider a proposal distribution that modifies one or more of the
components $\{\xx_i\}$, while keeping the number of components $N$, as well as the
hyperparameters $\hyperparams$, fixed.

@@

Choosing just one object to change is most appropriate when the DNS
distribution is very constrained. When it is close to the prior, bolder
proposals that change more than one object at a time are possible. Hence,
we choose the number of components to change from a heavy tailed distribution
wherer the most probable value is 1 but there is also a nontrivial probability
of proposing to change $\sim N$ components at once.

\subsection{Proposals that change the hyperparameters,
keeping the components fixed}\label{sec:proposal3}
Another kind of proposal that we will consider is a proposal that keeps all of
the components fixed in place (i.e. leaves $\{\xx_i\}$) unchanged, but changes
the hyperparameter(s) from their current value $\hyperparams$
to a new value $\hyperparams'$. The proposal for the hyperparameter(s) is chosen
to satisfy detailed balance with respect to $p(\hyperparams)$ and should be
heavy-tailed.

The overall Metropolis acceptance probability
for this kind of move, if sampling the prior (or the constrained prior of DNS)
must include the following factor:
\begin{eqnarray}
\frac{\prod_{i=1}^N p(\xx_i | \hyperparams')}{\prod_{i=1}^N p(\xx_i | \hyperparams)}
\label{eqn:acceptance_prob}
\end{eqnarray}
Since this proposal leaves the components $\{\xx_i\}$ fixed, it will usually not
affect the value of the likelihood, and therefore the likelihood will not need
to be recomputed. Therefore this proposal is most useful for mixing the values
of the hyperparameters $\hyperparams$ when the target distribution is highly
constrained.

\subsection{Proposals that change the hyperparameters
and all of the components}\label{sec:proposal4}
The above proposals allow for changes to $N$, $\hyperparams$, and $\{\xx_i\}$,
and are therefore sufficient to allow for ``correct'' exploration of the
prior distribution, and the constrained prior distributions used in DNS.
However, they do not necessarily allow for {\it efficient} exploration, even
of the prior itself. The main reason for this is the inability for large
changes to be made to the hyperparameters $\hyperparams$. If the proposed change
from $\hyperparams$ to $\hyperparams'$ is large, the ratio in
Equation~\ref{eqn:acceptance_prob} is likely to be very small, and the move
will probably be rejected.

Therefore, we allow an additional move that changes $\hyperparams$ to a new
value $\hyperparams'$, but rather than leaving the components $\{\xx_i\}$ fixed,
they are ``dragged''
so they represent the distribution $p(\xx | \hyperparams')$ rather than
$p(\xx|\hyperparams)$. We do this by making use of the transformation functions
$F(x; \hyperparams)$ and $G(x; \hyperparams)$ defined in
Section~\ref{sec:proposal2}.

The ``dragging'' process works as follows, and must be carried out on
each object:
\begin{eqnarray}
u_i &:=& F(\xx_i; \hyperparams)\\
\xx_i' &:=& G(\xx_i; \hyperparams')
\end{eqnarray}
In other words, the components' parameters are transformed to $[0,1]$ using the
current value of the hyperparameters, and then transformed back using the
proposed value of the hyperparameters so they represent the new interim prior
rather than the old one.

@@

%\section{``Transit'' Example}
%\section{``Asteroseismology'' Example}
\section{``Galaxy Field'' Example}
Source detection is an important problem in astrophysics. Given some data,
usually one or more noisy, blurred, images of the sky, we would like to know
how many components $N$ (such as stars or galaxies) are in the image, and their
properties (such as flux, size, orientation). Various approaches exist,
covering a wide spectrum between ad-hoc approaches and principled inference
approaches
\citep[e.g.][]{irwin, sextractor, dolphot, 2003MNRAS.338..765H, starfield}.
The more principled techniques tend to be much more computationally
intensive, so should only be applied to interesting or
especially challenging subsets of astronomical imaging. In fact, the approach
used by \citet{starfield} was essentially an early version of the methodology
presented in this paper.

In this section we apply
the DNS approach to a toy version of the problem of detecting and quantifying
extended components such as galaxies in a noisy image.
We model each galaxy as a mixture of two concentric elliptical gaussian
components, with
the following eight parameters. Firstly, two
parameters $x_c$ and $y_c$ describe the central position of the galaxy within
the image. The flux $f$ describes the total integral of the galaxy's
intensity profile. The axis ratio $q$ describes the ellipticity of the galaxy
and $\theta$ its orientation angle with respect to horizontal. The radius of the
bigger gaussian is a parameter $w$.
Finally
we include a ``radius ratio'' $u$ describing the ratio of the radius of the
smaller gaussian with respect to that of the bigger gaussian, and a parameter
$v$ describing the fraction of the total flux in the smaller gaussian
(therefore the fraction of the light in the bigger gaussian is $1-v$).
In a coordinate system aligned with the major axis of the ellipse, the
surface brightness profile of a ``galaxy'' is given by
\begin{eqnarray}
\rho(x, y) &=&
\frac{f(1-v)}{2\pi w^2}\exp\left[-\frac{1}{2w^2}\left(qx^2 + y^2/q\right)\right]\\
&&+\frac{fv}{2\pi (uw)^2}\exp\left[-\frac{1}{2(uw)^2}\left(qx^2 + y^2/q\right)\right].
\end{eqnarray}
We generated a simulated $200 \times 200$ pixel noisy image
(Figure~\ref{fig:galaxyfield_data}) containing 47 ``galaxies'',
in order to test the algorithm.

@@

However, many of the proposal distributions used involve changing a subset
of the parameters, while keeping others fixed. Considerable speedups can be
achieved by, when appropriate, updating the mock signal to reflect the proposed
change to the model, rather than computing the entire mock signal from scratch.
For example, if the proposal is to add two new components to the model, the
mock signal can simply be updated by adding the effect of the two new components
to the model.
